Different forms of uncertainty differently affect the evolution of social learning

### 2022-07-15 ms.tex
updated nagents sensitivity and started Model revision

To answer our research questions we developed an agent-based model of social
learning evolution in which a society of $N$ individuals \cm{do we ever define N used?} each must decide which of
$B$ behaviors to perform at each time step within a generation of $L$ time steps. At
the end of each generation, agents are selected with replacement from the population
to reproduce another $N$ agents. Agents who inherit the social learning trait then
learn about behavioral payoffs from the previous generation, while asocial learners
begin life with no \emph{a priori} knowledge of the world.  Each behavior is a
``bandit'', a common modeling and experimental approach for representing behaviors
with probabilistic
payoffs~\cite{SuttonBartoBook,McElreath2005,Steyvers2009,Rendell2010,Schulz2019}.  Each behavior
$b$ yields Bernoulli payoffs: payoff of 1 with probability $\pi_b$ and zero payoff
with probability $1 - \pi_b$; $\pi_b$ is therefore also the expected payoff of
behavior $b$. One behavior in each generation is optimal and pays off $\pihigh$,
while the rest pay off $\pilow < \pihigh$.  Model agents must decide which behavior
to perform at each time step.  To do this, agents use an explore-exploit strategy to
sometimes try the most profitable behavior they know about, and other times try
alternatives that may pay off more reliably.  We developed a series of computational
analyses by systematically varying uncertainty parameters and observing the
frequency that model populations evolve to be social learners, the average payoffs
accrued by agents in each setting, and the time it takes for model populations to
reach fixation (agents all social or all asocial learners). Model dynamics are
illustrated in Figure~\ref{fig:schematic}.

@@

        $\tau$ & Softmax temp.; $\uparrow=$more exploration, $\downarrow=$more
                    exploitation 
               & 0.01, \textbf{0.1}, 1.0 \\

@@

@@

At generation start, one of the behaviors is set to 
yield expected payoff $\pihigh$, with the other $B-1$ behaviors yielding
expected payoffs $\pilow$. At each time step, agents perform behavior $b$ 
with softmax-weighted probability
\begin{equation}
  \Pr(\text{Agent $i$ performs behavior $b$}) = 
    \frac{e^{\beta \ledger}}{\sum_{b=1}^B e^{\beta \ledger}},
\end{equation}
\noindent
where $\beta$ is a parameter that adjusts how frequently agents perform 
behaviors with high expected payoffs (larger $\beta$) versus how frequently
agents explore alternative behaviors (smaller $\beta$)~\cite{McElreath2005}. 
Agents probabilistically recieve a payoff of 0 or 1 depending on whether the
bandit paid off, equivalent to a draw from a Bernoulli distribution with 
mean $\pi_b$. For concreteness we denote this probabilistic payoff
$\bandit$. On performing behavior $b$, agent $i$ updates the
corresponding behavior count by 1, $c_{ib} \leftarrow c_{ib} + 1$, and then
the expected payoffs calculated for that behavior are updated with
exponentially-weighted averaging
\begin{equation}
  \ledger \leftarrow \ledger + \frac{\bandit - \ledger}{c_{ib}}.
\end{equation}
\noindent
This procedure continues for $L$ time steps within each generation
(Figure~\ref{fig:schematic}B).

@@

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied their values and observed three key outcome variables aggregated across 1000
trial simulations, and across all agents when appropriate. We varied $u \in \{0.0,
0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L
\in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  We observed
three outcome measures: (1) $\meansl$, the average value of $s_i$ over all agents
and trials; (2) $\meanpi / L$, the average net payoffs at simulation end across
agents and trials, normalized by lifespan; and (3) $\meanT / L$, 
or the mean number of time steps to fixation normalized by
lifespan\footnote{Equivalent to the mean number of generations to fixation}
(fixation means $\sum_i s_i = 0 \text{ or } 1$). \mt{Make sure to qualitatively
introduce these and their evolutionary importance in the intro.}
We analyze outcomes by plotting $\meansl$ on the y-axis and environmental
variation is on x-axis since we theoretically expect that $\meansl$ will 
decrease monotonically from 1 to 0 as $u$ increases. We expected and observed
that the other three uncertainty factors shift the value of $u$ at which 
$\meansl$ starts to decrease, and how rapidly $\meansl$ decreases. Note that the
hypothesis-testing concept of significance is meaningless here because we could
make any small outcome difference ``significant'' by running more simulation trials.
We instead study the patterns of outcome 
variables over systematically varied uncertainty factor values. 
Our model was implemented in the Julia programming language~\cite{Bezanson2017} 
using the Agents.jl agent-based modeling library~\cite{Datseris2022} and run
on the Sherlock supercomputing cluster at Stanford University. Model code and
data are publicly available on GitHub\footnote{\url{https://github.com/mt-digital/UncMod}} 
and OSF.io\footnote{\mt{TODO}}.

>>>>

To answer our research questions we developed an agent-based model of social
learning evolution in which a society of $N$ individuals \cm{do we ever define N used?} each must decide which of
$B$ behaviors to perform at each time step within a generation of $L$ time steps. At
the end of each generation, agents are selected with replacement from the population
to reproduce another $N$ agents. Agents who inherit the social learning trait then
learn about behavioral payoffs from the previous generation, while asocial learners
begin life with no \emph{a priori} knowledge of the world.  Each behavior is a
``bandit'', a common modeling and experimental approach for representing behaviors
with probabilistic
payoffs~\cite{SuttonBartoBook,McElreath2005,Steyvers2009,Rendell2010,Schulz2019}.

@@

        $\beta$ & Softmax greediness; $\uparrow=$more exploitation, $\downarrow=$more
                    exploration 
               & 100, \textbf{10}, 1 \\

@@

\subsection{Agents and learning}

All agents perform individual, trial-and-error learning over their lifespan.
This learning is guided based on undirected, softmax search which drives
agents to explore all behavioral options in the selection set when 
the agent is more uncertain which is optimal, and to exploit the most profitable
behaviors when the agent is more certain which are most beneficial.

\subsection{Dynamics}

\subsubsection{Initialization}

At generation start, one of the behaviors is set to 
yield expected payoff $\pihigh$, with the other $B-1$ behaviors yielding
expected payoffs $\pilow$. At each time step, agents perform behavior $b$ 
with softmax-weighted probability
\begin{equation}
  \Pr(\text{Agent $i$ performs behavior $b$}) = 
    \frac{e^{\beta \ledger}}{\sum_{b=1}^B e^{\beta \ledger}},
\end{equation}
\noindent
where $\beta$ is a parameter that adjusts how frequently agents perform 
behaviors with high expected payoffs (larger $\beta$) versus how frequently
agents explore alternative behaviors (smaller $\beta$)~\cite{McElreath2005}. 
Agents probabilistically recieve a payoff of 0 or 1 depending on whether the
bandit paid off. Behavior payoffs are represented as a draw from a 
Bernoulli distribution with mean $\pi_b$, so $\pi_b$ is also the mean payoff from
behavior $b$.  For concreteness we denote this probabilistic payoff
$\bandit$. On performing behavior $b$, agent $i$ updates the
corresponding \emph{behavior count} by 1, $c_{ib} \leftarrow c_{ib} + 1$, and then
the expected payoffs calculated for that behavior are updated with
exponentially-weighted averaging
\begin{equation}
  \ledger \leftarrow \ledger + \frac{\bandit - \ledger}{c_{ib}}.
\end{equation}
\noindent
This procedure continues for $L$ time steps within each generation
(Figure~\ref{fig:schematic}B).

\subsubsection{Intragenerational dynamics}

@@

\subsubsection{Intergenerational dynamics}

@@

\subsection{Computational analyses}

We developed a series of computational
analyses by systematically varying uncertainty parameters and observing the
frequency that model populations evolve to be social learners, the average payoffs
accrued by agents in each setting, and the time it takes for model populations to
reach fixation (agents all social or all asocial learners). Model dynamics are
illustrated in Figure~\ref{fig:schematic}.

@@

To analyze the effect of the four principal uncertainty factors, we systematically
varied their values and observed three key outcome variables aggregated across 1000
trial simulations, and across all agents when appropriate. We varied $u \in \{0.0,
0.1, \ldots, 1.0\}$; $\pilow \in \{0.1, 0.45, 0.8\}$; $B \in \{2, 4, 10\}$; and $L
\in \{1,2,4,8\}$ for $B=2$ and $L \in \{1,B/2,B,2B\}$ for $B=4,10$.  We observed
three outcome measures: (1) $\meansl$, the average value of $s_i$ over all agents
and trials; (2) $\meanpi / L$, the average net payoffs at simulation end across
agents and trials, normalized by lifespan; and (3) $\meanT / L$, 
or the mean number of time steps to fixation normalized by
lifespan\footnote{Equivalent to the mean number of generations to fixation}
(fixation means $\sum_i s_i = 0 \text{ or } 1$). \mt{Make sure to qualitatively
introduce these and their evolutionary importance in the intro.}
We analyze outcomes by plotting $\meansl$ on the y-axis and environmental
variation is on x-axis since we theoretically expect that $\meansl$ will 
decrease monotonically from 1 to 0 as $u$ increases. We expected and observed
that the other three uncertainty factors shift the value of $u$ at which 
$\meansl$ starts to decrease, and how rapidly $\meansl$ decreases. Note that the
hypothesis-testing concept of significance is meaningless here because we could
make any small outcome difference ``significant'' by running more simulation trials.
We instead study the patterns of outcome 
variables over systematically varied uncertainty factor values.

\subsubsection{Implementation}
Our model was implemented in the Julia programming language~\cite{Bezanson2017} 
using the Agents.jl agent-based modeling library~\cite{Datseris2022} and run
on the Sherlock supercomputing cluster at Stanford University. Model code and
data are publicly available on GitHub\footnote{\url{https://github.com/mt-digital/UncMod}}.
